---
title: "Darwin Core mapping"
subtitle: "For: Checklist of alien mammals in Belgium"
author:
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
---

This document describes how we map the checklist data to Darwin Core. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
```

# Read source data

The data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1lqGylVY4SL4AuW3mKxEmHYXEFppUMaM-FLplKzAJ2t0/edit#gid=0).

Read the relevant worksheet (published as csv):

```{r read_source_data}
input_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTpIbZpKCr25l-1PuUuSP8Jk191uwZtTXy5HvXJ7P8ZCPNDitkicpQhwKmnfymeo9ERV2n6pbsrOWQG/pub?gid=0&single=true&output=csv")
```

Copy the source data to the repository to keep track of changes:

```{r}
write_csv(input_data, here("data", "raw", "alien_mammals_checklist_dump.csv"), na = "")
```

Preview data: 

```{r}
input_data %>% head()
```

# Preprocessing

## Tidy data

Clean data somewhat:

```{r}
input_data %<>%
  remove_empty("rows") %>%    # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Scientific names

No cleaning required.

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom (that will remain the same as long as scientific name and kingdom remain the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "alien-mammals-checklist",
  "taxon",
  vdigest(paste(scientific_name, kingdom), algo = "md5"),
  sep = ":"
))
```


Preview data:

```{r}
input_data %>% head()
```

# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
taxon <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "INBO") 
```

### accessRights

```{r}
taxon %<>% mutate(dwc_accessRights = "https://www.inbo.be/en/norms-data-use") 
```

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "") 
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "INBO") 
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "Checklist of alien mammals of Belgium") 
```

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name) 
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = kingdom)
```

### phylum

```{r}
taxon %<>% mutate(dwc_phylum = phylum)
```

### order

```{r}
taxon %<>% mutate(dwc_order = order)
```

### family

```{r}
taxon %<>% mutate(dwc_family = family)
```

### genus

```{r}
taxon %<>% mutate(dwc_genus = genus)
```

### taxonRank

```{r}
taxon %<>% mutate(dwc_taxonRank = taxon_rank)
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = nomenclatural_code)
```