---
title: "Darwin Core mapping"
subtitle: "For: Checklist of alien mammals in Belgium"
authors:
- Lien Reyserhove
- Sanne Govaert
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_depth: 3
    toc_float: true
---

This document describes how we map the checklist data to Darwin Core. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
```

# Read source data

The data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1lqGylVY4SL4AuW3mKxEmHYXEFppUMaM-FLplKzAJ2t0/edit#gid=0).

Read the relevant worksheet (published as csv):

```{r read_source_data}
input_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTpIbZpKCr25l-1PuUuSP8Jk191uwZtTXy5HvXJ7P8ZCPNDitkicpQhwKmnfymeo9ERV2n6pbsrOWQG/pub?gid=0&single=true&output=csv", col_types = cols(.default = "c"))
```

Copy the source data to the repository to keep track of changes:

```{r}
write_csv(input_data, here("data", "raw", "alien_mammals_checklist_dump.csv"), na = "")
```

Preview data: 

```{r}
input_data %>% head()
```

# Preprocessing

## Tidy data

Clean data somewhat:

```{r}
input_data %<>%
  remove_empty("rows") %>%    # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Scientific names

No cleaning required.

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom (that will remain the same as long as scientific name and kingdom remain the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "alien-mammals-checklist",
  "taxon",
  vdigest(paste(scientific_name, kingdom), algo = "md5"),
  sep = ":"
))
```


Preview data:

```{r}
input_data %>% head()
```

# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows). Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

```{r}
taxon <-
  input_data %>%
  distinct(taxon_id, .keep_all = TRUE) %>%
  mutate(
    dwc_language = "en",
    dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/",
    dwc_rightsHolder = "INBO",
    dwc_accessRights = "https://www.inbo.be/en/norms-data-use",
    dwc_datasetID = "https://doi.org/10.1016/j.tree.2011.03.023",
    dwc_institutionCode = "INBO",
    dwc_datasetName = "Checklist of alien mammals of Belgium",
    dwc_taxonID = taxon_id,
    dwc_scientificName = scientific_name,
    dwc_kingdom = kingdom,
    dwc_phylum = phylum,
    dwc_order = order,
    dwc_family = family,
    dwc_genus = genus,
    dwc_taxonRank = taxon_rank,
    dwc_nomenclaturalCode = nomenclatural_code
  )
```

## Distribution extension

Create a dataframe with all data (including multiple distributions). Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

Information for `eventDate` is contained in `date_first_observation` and `date_last_observation`, which we will express here in an ISO 8601 date format `yyyy/yyyy` (`start_date/end_date`).

Inspect `data_first_observation` and `date_last_observation`:

```{r}
input_data %>%
  group_by(date_first_observation, date_last_observation) %>%
  count()
```

Not all cells for `date_first_observation` and/or `date_last_observation` are populated. So, we used the following rules (`YP` = year of (re)publication of the checklist) for those records: 

***case 1.*** If `DFO` is empty and `DLO` is empty, `eventDate` is `YP/YP` 
***case 2.***  If `DFO` is empty and `DLO` is not empty: take `date_last_observation` as `date_first_observation`, eventDate = `DLO/DLO`
***case 3.*** If `DFO` is not empty and `DLO` is empty, take `YP` as `DLO`, so eventDate is `DFO/YP`

```{r}
# define current year
year_publication <- format(Sys.Date(), "%Y")

# dwc mapping
distribution <-
  input_data %>%
  mutate(
    dwc_taxonID = taxon_id,
    dwc_locationID = case_when(
      location == "Flanders" ~ "ISO_3166-2:BE-VLG",
      location == "Wallonia" ~ "ISO_3166-2:BE-WAL",
      location == "Brussels" ~ "ISO_3166-2:BE-BRU"
    ),
    dwc_locality = case_when(
      location == "Flanders" ~ "Flemish Region",
      location == "Wallonia" ~ "Walloon Region",
      location == "Brussels" ~ "Brussels-Capital Region"
    ),
    dwc_countryCode = "BE",
    dwc_occurrenceStatus = occurrence_status,
    dwc_establishmentMeans = "introduced",
    dwc_degreeOfEstablishment = degree_of_establishment,
    dwc_eventDate = case_when(
      is.na(date_first_observation) & is.na(date_last_observation) ~ paste(year_publication, year_publication, sep = "/"),
      is.na(date_first_observation) & !is.na(date_last_observation) ~ paste(date_last_observation, date_last_observation, sep = "/"),
      !is.na(date_first_observation) & is.na(date_last_observation) ~ paste(date_first_observation, year_publication, sep = "/"),
      !is.na(date_first_observation) & !is.na(date_last_observation) ~ paste(date_first_observation, date_last_observation, sep = "/")
    ),
    dwc_source = source,
    dwc_occurrenceRemarks = remarks
  )
```

Check mapping of eventDate:

```{r}
distribution %>%
  group_by(date_first_observation, date_last_observation, dwc_eventDate) %>%
  count()
```

## Species profile extension

In this extension we will express broad habitat characteristics of the species (e.g. `isTerrestrial`) from `realm`.

Create a dataframe with unique taxa only (ignoring multiple distribution rows).
Only keep records for which `realm` is not empty.

```{r}
species_profile <-
  input_data %>%
  distinct(taxon_id, .keep_all = TRUE) %>% 
  filter(!is.na(realm))
```

Inspect values:

```{r}
species_profile %>%
  group_by(realm) %>%
  count()
```

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

### taxonID

```{r}
species_profile %<>%
  mutate(
    dwc_taxonID = taxon_id,
    dwc_isMarine = "FALSE",
    dwc_isFreshwater =
      case_when(
        realm == "freshwater" ~ "TRUE",
        TRUE ~ "FALSE"
      ),
    dwc_isTerrestrial =
      case_when(
        realm == "terrestrial" ~ "TRUE",
        TRUE ~ "FALSE"
      )
  )
```

Show mapped values:

```{r}
species_profile %>%
  select(realm, dwc_isMarine, dwc_isFreshwater, dwc_isTerrestrial) %>%
  group_by_all() %>%
  summarize(records = n())
```
## Description extension

In the description extension we want to include the species' native range. We will express native range information from `native_range`, separated and gathered.

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
description <-
  input_data %>%
  distinct(taxon_id, .keep_all = TRUE) %>%
  # Separate values on `|` in a maximum of 4 columns:
  separate(
    native_range,
    into = paste0("range_", c(1:4)),
    sep = " \\| ",
    extra = "drop"
  ) %>%
  # Gather and trim values:
  gather(key, value, starts_with("range_"), na.rm = TRUE) %>%
  mutate(value = str_trim(value))
```

Inspect values:

```{r}
description %>%
  group_by(value) %>%
  count()
```

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

```{r}
description %<>%
  mutate(
    dwc_taxonID = taxon_id,
    dwc_description = value,
    dwc_type = "native range",dwc_language = "en")
```

# Post-processing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
distribution %<>% select(starts_with("dwc_"))
species_profile %<>% select(starts_with("dwc_"))
description %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
colnames(species_profile) <- str_remove(colnames(species_profile), "dwc_")
colnames(description) <- str_remove(colnames(description), "dwc_")
```

Sort on `taxonID` (to maintain some consistency between updates of the dataset):

```{r}
taxon %<>% arrange(taxonID)
distribution %<>% arrange(taxonID)
species_profile %<>% arrange(taxonID)
description %<>% arrange(taxonID)
```

Preview taxon core:

```{r}
taxon %>% head()
```
Preview distribution extension:

```{r}
distribution %>% head()
```

Preview species profile extension:

```{r}
species_profile %>% head()
```

Preview description extension:

```{r}
description %>% head(10)
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```

